---
title: "Bayesian Inference for Lineup Measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Inference for Lineup Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Purpose

This vignette shows the Bayesian posterior summaries in **r4lineups** for
lineup fairness and confidence-based eyewitness measures.

Use these functions when you want credible intervals or direct probability
statements such as \(P(E' < 6 \mid data)\), \(P(F < 6 \mid data)\), or
\(P(DR > 1 \mid data)\). The examples use small posterior samples so that the
vignette runs quickly during package checks; increase `S` for final analyses.

```{r}
library(r4lineups)
```

# Data Shapes

The Bayesian fairness functions use mock-witness lineup choices:

- a **lineup vector**, where each value is the chosen lineup position; or
- a **lineup table**, where each cell is the count for one lineup position.

The Bayesian calibration function uses the standard confidence-based data frame
with `target_present`, `identification`, and `confidence`. The SDT comparison
uses summary counts: hits, misses, false alarms, and correct rejections for two
conditions.

# Tredoux Effective Size

`esize_T_bayes()` fits a Dirichlet-multinomial model to the lineup choice counts
and transforms each posterior draw of the choice-probability vector to Tredoux's
effective size \(E'\).

```{r}
data(nortje2012)

lineup_vec <- nortje2012$lineup_1
k <- max(lineup_vec, na.rm = TRUE)
lineup_table <- table(factor(lineup_vec, levels = seq_len(k)))

set.seed(1)
esize_post <- esize_T_bayes(
  lineup_table,
  alpha = 0.5,
  S = 1000,
  threshold = k
)

esize_post
```

```{r}
plot(esize_post)
```

The threshold argument makes the posterior useful for fairness benchmarks. For
example, with a six-person lineup one might inspect \(P(E' < 6 \mid data)\) as a
posterior probability that the effective size falls below nominal size.

# Functional Size

`func_size_bayes()` models the suspect-selection rate with a beta-binomial
posterior and transforms posterior draws to functional size.

```{r}
set.seed(2)
func_post <- func_size_bayes(
  lineup_vec,
  target_pos = 3,
  alpha = 0.5,
  S = 1000,
  threshold = k
)

func_post
```

```{r}
plot(func_post)
```

# Diagnosticity Ratio

`diag_ratio_T_bayes()` models target-present and target-absent suspect-selection
rates with independent beta-binomial posteriors and returns the posterior of the
diagnosticity ratio.

```{r}
tp_vec <- nortje2012$lineup_1
ta_vec <- nortje2012$lineup_2

set.seed(3)
diag_post <- diag_ratio_T_bayes(
  lineup_pres = tp_vec,
  lineup_abs = ta_vec,
  pos_pres = 3,
  pos_abs = 3,
  k1 = max(tp_vec, na.rm = TRUE),
  k2 = max(ta_vec, na.rm = TRUE),
  alpha = 0.5,
  S = 1000,
  threshold = 1
)

diag_post
```

```{r}
plot(diag_post)
```

The threshold `1` is useful because \(DR > 1\) indicates that suspect
identifications are more common in the target-present lineup than in the
target-absent lineup.

# Calibration

`calibration_bayes()` places beta posteriors on accuracy within confidence bins
and propagates uncertainty into calibration \(C\), over/underconfidence
\(O/U\), and normalized resolution \(NRI\).

```{r}
data(lineup_example)

set.seed(4)
cal_post <- calibration_bayes(
  lineup_example,
  confidence_bins = c(0, 60, 80, 100),
  alpha = 0.5,
  S = 1000
)

cal_post
```

```{r}
plot(cal_post, metric = "C")
plot(cal_post, metric = "NRI")
```

# SDT d' Comparison

`sdt_compare()` compares two conditions using summary SDT counts. It draws from
posterior hit and false-alarm rates, converts each draw to \(d'\), and reports
the posterior for \(\Delta d' = d'_A - d'_B\).

```{r}
set.seed(5)
sdt_post <- sdt_compare(
  hits_A = 45, misses_A = 55, fas_A = 12, crs_A = 88,
  hits_B = 38, misses_B = 62, fas_B = 10, crs_B = 90,
  label_A = "Procedure A",
  label_B = "Procedure B",
  alpha = 0.5,
  S = 1000
)

sdt_post
```

```{r}
plot(sdt_post)
```

# Workflow Position

These Bayesian functions complement, rather than replace, the frequentist and
bootstrap functions elsewhere in the package:

- use normal-theory functions for quick classical intervals where assumptions are
  acceptable;
- use bootstrap functions when you want sampling-distribution uncertainty with
  minimal distributional assumptions;
- use Bayesian functions when credible intervals and direct posterior
  probabilities are the clearest way to express uncertainty.

For substantive reports, run enough posterior draws for stable summaries and
check sensitivity to the prior concentration, for example `alpha = 0.1`, `0.5`,
and `1`.
