---
title: "Calibration and Decision Analysis"
subtitle: "Advanced Methods for Evaluating Eyewitness Identification Performance"
author:
- "Colin G. Tredoux"
- "Tamsyn M. Naylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration and Decision Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates five advanced statistical methods for analyzing eyewitness identification data in **r4lineups**:

1. **Calibration Analysis** - Assessing confidence-accuracy correspondence
2. **Bayesian Information-Gain Curves** - Evaluating diagnostic value of evidence
3. **Expected Utility Analysis** - Decision-theoretic evaluation with cost/benefit tradeoffs
4. **Deviation from Perfect Performance (DPP)** - ROC-based truncation-robust metric
5. **ANRI** - Bias-corrected resolution index with bootstrap inference

These methods complement the traditional lineup fairness measures by providing sophisticated analyses of identification accuracy, confidence calibration, and decision-making utility.

## Data Format

All methods in this vignette require a dataframe with the following columns:

- `target_present`: Logical. TRUE if the guilty suspect is in the lineup
- `identification`: Character. "suspect", "filler", or "reject"
- `confidence`: Numeric. Confidence rating (any scale)

We'll use the built-in `lineup_example` dataset throughout.

```{r load-data}
library(r4lineups)
data(lineup_example)

# Examine the data structure
head(lineup_example)
str(lineup_example)
```

## 1. Calibration Analysis

Calibration measures how well confidence judgments correspond to accuracy. Perfect calibration means that when witnesses express X% confidence, they are correct X% of the time.

### Basic Calibration

```{r calibration-basic}
# Compute calibration with binned confidence
cal_result <- make_calibration(
  lineup_example,
  confidence_bins = c(0, 60, 80, 100),
  choosers_only = TRUE  # Analyze only suspect identifications
)

print(cal_result)
```

The output provides three key statistics:

- **C statistic**: Overall calibration quality (0 = perfect, higher = worse)
- **O/U (Over/Under confidence)**: Positive = overconfident, Negative = underconfident
- **NRI (Normalized Resolution Index)**: Ability to discriminate accuracy levels with confidence (higher = better)

### Interpreting the Calibration Plot

The calibration plot shows:

- **Diagonal line**: Perfect calibration
- **Points above diagonal**: Underconfident (accuracy exceeds confidence)
- **Points below diagonal**: Overconfident (confidence exceeds accuracy)
- **Point size**: Sample size in each bin

```{r calibration-plot, eval=FALSE}
# The plot is automatically displayed, but you can save it:
ggsave("calibration_plot.png", cal_result$plot, width = 8, height = 6)
```

### Calibration by Condition

Compare calibration across experimental conditions:

```{r calibration-by-condition}
# Add a grouping variable (for demonstration)
lineup_example$procedure <- sample(
  c("Simultaneous", "Sequential"),
  nrow(lineup_example),
  replace = TRUE
)

# Compute calibration by condition
cal_by_cond <- make_calibration_by_condition(
  lineup_example,
  condition_var = "procedure",
  confidence_bins = c(0, 60, 80, 100)
)

print(cal_by_cond)

# Plot comparison
cal_by_cond$plot
```

## 2. Bayesian Information-Gain Analysis

Bayesian analysis evaluates how much an identification response changes our belief about guilt, quantified as information gain (reduction in uncertainty).

### Prior-Posterior Curves

```{r bayes-curves}
# Compute Bayesian curves for simple responses
bayes_result <- make_bayes_curves(
  lineup_example,
  response_categories = "simple",  # "suspect", "filler", "reject"
  prior_grid = seq(0.01, 0.99, 0.01)
)

print(bayes_result)

# Plot prior-posterior relationships
plot_bayes_prior_posterior(bayes_result)
```

The plot shows how each response type updates prior belief:

- **Lines above diagonal**: Evidence increases guilt belief (diagnostic of guilt)
- **Lines below diagonal**: Evidence decreases guilt belief (diagnostic of innocence)
- **Steeper slopes**: More diagnostic evidence

### Information Gain

Information gain quantifies uncertainty reduction:

```{r bayes-information-gain}
# Plot information gain
plot_bayes_information_gain(bayes_result)
```

Interpretation:

- **Positive values**: Response reduces uncertainty (diagnostic)
- **Negative values**: Response increases uncertainty (misleading)
- **Higher magnitude**: Stronger diagnostic value

### Confidence-Based Bayesian Analysis

Analyze by confidence level:

```{r bayes-confidence}
bayes_conf <- make_bayes_curves(
  lineup_example,
  response_categories = "confidence",
  confidence_bins = c(0, 60, 80, 100)
)

# Get suspect responses only
suspect_responses <- grep("^suspect_", unique(bayes_conf$curves$response), value = TRUE)
plot_bayes_prior_posterior(bayes_conf, selected_responses = suspect_responses)
```

### Base-Rate Equivalency of Evidence (BREE)

BREE curves compare the diagnostic value of two procedures:

```{r bree}
# Create two procedure datasets (for demonstration)
data_proc_a <- lineup_example[lineup_example$procedure == "Simultaneous", ]
data_proc_b <- lineup_example[lineup_example$procedure == "Sequential", ]

# Compute BREE curve for suspect identifications
bree_result <- make_bree_curve(
  data_proc_a,
  data_proc_b,
  reference_response = "suspect"
)

plot_bree(bree_result)
```

The BREE curve shows equivalent base rates for equal posterior probabilities. Points above the diagonal indicate Procedure A requires higher base rates to achieve the same posterior as Procedure B.

## 3. Expected Utility Analysis

Expected utility analysis evaluates identification procedures considering costs and benefits of different outcomes.

### Define Utility Matrix

```{r utility-matrix}
# Define costs/benefits (on same scale)
utility_matrix <- c(
  tp = 1.0,    # True positive (correct suspect ID): Benefit
  fn = -0.5,   # False negative (reject/filler ID when guilty): Cost
  fp = -2.0,   # False positive (suspect ID when innocent): Major cost
  tn = 0.5     # True negative (reject/filler ID when innocent): Benefit
)
```

### Compute Utility Curves

```{r utility-curves}
# Compute expected utility across confidence criteria
util_result <- make_utility_curves(
  lineup_example,
  base_rate = 0.5,  # Assume 50% guilty suspects
  utility_matrix = utility_matrix,
  lineup_size = 6
)

print(util_result)

# Plot utility curves
plot_utility_curves(util_result)
```

The plot shows:

- **Expected utility** at each confidence threshold
- **Optimal threshold**: Maximum utility point
- **Comparison with "all identifications"** strategy

### Compare Procedures

```{r utility-comparison, eval=FALSE}
# Compare two procedures (requires datasets from BREE section)
# First compute utility for each procedure separately
util_proc_a <- make_utility_curves(data_proc_a, base_rate = 0.5,
                                   utility_matrix = utility_matrix)
util_proc_b <- make_utility_curves(data_proc_b, base_rate = 0.5,
                                   utility_matrix = utility_matrix)

cat("Procedure A max utility:", round(util_proc_a$max_utility, 3), "\n")
cat("Procedure B max utility:", round(util_proc_b$max_utility, 3), "\n")
```

### Sensitivity to Base Rate

Examine how utility depends on base rate:

```{r utility-base-rates, eval=FALSE}
# Test multiple base rates
base_rates <- c(0.2, 0.5, 0.8)

for (br in base_rates) {
  util <- make_utility_curves(lineup_example, base_rate = br,
                              utility_matrix = utility_matrix)
  cat("Base rate:", br, "  Max utility:", round(util$max_utility, 3), "\n")
}
```

## 4. Deviation from Perfect Performance (DPP)

DPP quantifies how far an ROC curve deviates from perfect performance. It's robust to ROC truncation (missing low-confidence data).

### Basic DPP

```{r dpp}
# Compute DPP
dpp_result <- make_dpp(
  lineup_example,
  lineup_size = 6
)

print(dpp_result)

# Plot ROC with perfect performance comparison
plot_dpp(dpp_result)
```

Interpretation:

- **DPP = 0**: Perfect performance
- **DPP = 1**: Chance performance
- **DPP < 0**: Better than perfect (impossible, indicates error)

The plot shows:

- **Black curve**: Observed ROC
- **Red dashed line**: Perfect ROC given the maximum false alarm rate
- **Shaded area**: Deviation from perfect performance

### Compare DPP Between Groups

```{r dpp-comparison}
# Compare procedures (using datasets from BREE section)
dpp_comparison <- compare_dpp(
  data_proc_a,
  data_proc_b,
  lineup_size = 6
)

print(dpp_comparison)

# Plot comparison side-by-side
plot_dpp_comparison(dpp_comparison)
```

### When to Use DPP

Use DPP when:

- You have confidence-based ROC data
- ROC curves are truncated (witnesses may not use full scale)
- You want a single performance metric comparable across studies
- You need to compare procedures with different confidence distributions

## 5. ANRI: Adjusted Normalized Resolution Index

ANRI corrects NRI for small-sample bias, making it more appropriate for hypothesis testing and group comparisons.

### Basic ANRI

```{r anri}
# Compute ANRI
anri_result <- compute_anri(
  lineup_example,
  confidence_bins = c(0, 60, 80, 100)
)

print(anri_result)
```

ANRI is always ≤ NRI because it removes positive bias. The bias correction is larger when:

- Sample size N is small
- Number of bins J is large
- The ratio J/N is high

### Bootstrap Confidence Intervals

```{r anri-bootstrap}
# Compute ANRI with bootstrap CIs
anri_boot <- bootstrap_anri(
  lineup_example,
  confidence_bins = c(0, 60, 80, 100),
  n_bootstrap = 2000,  # Use 2000+ for publications
  conf_level = 0.95,
  seed = 123
)

print(anri_boot)

cat("\nANRI:", round(anri_boot$anri, 3), "\n")
cat("95% CI: [", round(anri_boot$ci_lower, 3), ",",
    round(anri_boot$ci_upper, 3), "]\n")
```

### Compare Groups

```{r anri-comparison}
# Compare ANRI between procedures
anri_comparison <- compare_anri(
  lineup_example,
  group_var = "procedure",
  confidence_bins = c(0, 60, 80, 100),
  n_bootstrap = 1000,
  seed = 456
)

print(anri_comparison)

# Plot comparison
plot_anri_comparison(anri_comparison)

# Plot bootstrap distribution
plot_anri_difference_distribution(anri_comparison)
```

The comparison provides:

- **Point estimates** for each group
- **Difference with CI**: Tests whether groups differ
- **Significance test**: CI excludes 0 = significant difference
- **Bootstrap distributions**: For diagnostic checks

### When to Use ANRI vs NRI

Use ANRI when:

- Sample size is modest (N < 200)
- Number of bins is small (J = 3-5)
- Comparing across studies with different bin counts
- Conducting formal statistical tests

Use NRI when:

- Sample size is very large (N > 500)
- You're only reporting descriptive statistics
- Comparing to older studies that reported NRI

## Integrating Multiple Methods

For a comprehensive analysis, combine these methods:

```{r comprehensive-analysis, eval=FALSE}
# 1. Calibration: Are witnesses well-calibrated?
cal <- make_calibration(data, confidence_bins = c(0, 60, 80, 100))

# 2. DPP: Overall discriminability
dpp <- make_dpp(data, lineup_size = 6)

# 3. ANRI: Resolution with inference
anri <- bootstrap_anri(data, confidence_bins = c(0, 60, 80, 100),
                       n_bootstrap = 2000)

# 4. Utility: Policy decisions
util <- make_utility_curves(data, base_rate = 0.5,
                            utility_matrix = utility_matrix)

# 5. Bayesian: Information value
bayes <- make_bayes_curves(data, response_categories = "confidence",
                           confidence_bins = c(0, 60, 80, 100))
```

## Reporting Results

### Example Write-Up

"We analyzed identification performance using calibration analysis, DPP, and ANRI with bootstrap inference. Witnesses showed moderate calibration (C = 0.12, O/U = -0.05), indicating slight overconfidence. The DPP was 0.34 (95% CI [0.28, 0.41]), indicating performance well above chance but below perfection. Resolution, measured by ANRI, was 0.25 (95% CI [0.15, 0.36]), demonstrating that confidence ratings provided meaningful discrimination between correct and incorrect identifications.

Expected utility analysis revealed that a confidence threshold of 75% maximized utility (EU = 0.42) given a base rate of 50% and the specified cost matrix. Bayesian analysis showed that high-confidence suspect identifications provided the most information gain (mean IG = 0.48 bits), substantially reducing uncertainty about guilt."

## References

**Calibration:**
- Juslin, P., Olsson, N., & Winman, A. (1996). Calibration and diagnosticity of confidence in eyewitness identification. *Journal of Experimental Psychology: Learning, Memory, and Cognition, 22*(5), 1304-1316.

**Bayesian Analysis:**
- Wells, G. L., Yang, Y., & Smalarz, L. (2015). Eyewitness identification: Bayesian information gain, base-rate effect-equivalency curves, and reasonable suspicion. *Law and Human Behavior, 39*(2), 99-122.

**Expected Utility:**
- Lampinen, J. M., Smith, A. M., & Wells, G. L. (2019). Four utilities in eyewitness identification practice. *Journal of Applied Research in Memory and Cognition, 8*(1), 29-38.

**DPP:**
- Mickes, L. (2015). Receiver operating characteristic analysis and confidence-accuracy characteristic analysis in investigations of system variables and estimator variables that affect eyewitness memory. *Journal of Applied Research in Memory and Cognition, 4*(2), 93-102.

**ANRI:**
- Yaniv, I., Yates, J. F., & Smith, J. E. K. (1991). Measures of discrimination skill in probabilistic judgment. *Psychological Bulletin, 110*(3), 611-617.

## Session Info

```{r session-info}
sessionInfo()
```
