---
title: "Face Similarity Analysis"
subtitle: "Deep Learning for Objective Lineup Fairness Assessment"
author:
- "Colin G. Tredoux"
- "Tamsyn M. Naylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Face Similarity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5,
  eval = FALSE,
  purl = FALSE  # Requires local images and optional Python setup.
)
```

## Introduction

**r4lineups** provides deep learning methods for measuring facial similarity in eyewitness lineups. These tools complement mock-witness studies by giving analysts reproducible, continuous similarity summaries that can help screen candidate foils and document visual fairness checks. They should be treated as exploratory aids, not as a replacement for mock-witness or outcome-based evidence.

### Why Face Similarity?

Traditional lineup fairness assessment relies on mock witnesses viewing the lineup and making selections based on a description. While valuable, this approach:

- Requires recruiting and testing participants
- Is time-consuming and costly
- May be influenced by subjective factors
- Provides categorical data (selections) rather than continuous similarity scores

**Deep learning face similarity** offers:

- **Quantitative measurement** - Numeric similarity scores
- **Efficiency** - Analyze lineups in seconds
- **Scalability** - Process hundreds of lineups easily
- **Replicability** - Same lineup always gives same scores
- **Diagnostic detail** - Identify which specific foils are problematic

### What This Vignette Covers

1. **Setup** - Installing Python dependencies
2. **Basic Comparison** - Comparing two face images
3. **Lineup Analysis** - Analyzing entire lineups
4. **Models & Metrics** - Choosing the right configuration
5. **Visualization** - Displaying results
6. **Advanced Features** - Embeddings, batch processing, dimensionality reduction
7. **Integration** - Combining with traditional measures
8. **Practical Guidelines** - Interpreting similarity scores

## Setup

Face similarity analysis uses Python's DeepFace library via R's reticulate package. This requires a one-time setup.

### Check Current Setup

```{r check-setup, purl=FALSE}
library(r4lineups)

# Check if Python and dependencies are available
check_python_deps()
```

### Install Python Dependencies

If you don't have the required dependencies, install them:

```{r install-deps, purl=FALSE}
# This will:
# 1. Create a Python virtual environment
# 2. Install DeepFace and required packages
# 3. Download face detection models

install_r4lineups_python()
```

This may take several minutes the first time as it downloads deep learning models (ArcFace, FaceNet, etc.). The models are cached for future use.

### Verify Installation

```{r verify, purl=FALSE}
# Check available models
available_models()

# Check available face detectors
available_detectors()
```

## Basic Face Comparison

The simplest use case is comparing two face images.

### Compare Two Faces

```{r basic-comparison, purl=FALSE}
# Compare target to a foil
# (Use your own image paths)
result <- face_similarity(
  img1_path = "path/to/target.jpg",
  img2_path = "path/to/foil1.jpg",
  model = "ArcFace",     # State-of-the-art face recognition model
  metric = "cosine"      # Similarity metric
)

print(result)
```

### Understanding the Output

```{r interpret-result, purl=FALSE}
# Similarity score (0-1, higher = more similar)
result$similarity  # e.g., 0.85 = very similar

# Distance (lower = more similar)
result$distance    # e.g., 0.15

# Would faces be verified as same person?
result$verified    # TRUE = too similar for fair lineup

# Model's threshold for verification
result$threshold   # e.g., 0.68 for ArcFace + cosine
```

**Interpretation:**
- `similarity > 0.8`: Very similar, potential lineup bias
- `similarity 0.6-0.8`: Moderately similar
- `similarity < 0.6`: Sufficiently dissimilar
- `verified = TRUE`: Faces exceed model's same-person threshold

## Lineup Similarity Analysis

Analyze all lineup members against the target in one call.

### Analyze Complete Lineup

```{r lineup-analysis, purl=FALSE}
# Analyze all foils in a lineup
lineup_result <- lineup_similarity(
  target_path = "path/to/target.jpg",
  foil_paths = c(
    "path/to/foil1.jpg",
    "path/to/foil2.jpg",
    "path/to/foil3.jpg",
    "path/to/foil4.jpg",
    "path/to/foil5.jpg"
  ),
  model = "ArcFace",
  metric = "cosine"
)

print(lineup_result)
```

### Examining Results

```{r examine-lineup, purl=FALSE}
# View as dataframe
lineup_result

# Summary statistics
summary(lineup_result)

# Identify problematic foils (too similar)
problematic <- lineup_result[lineup_result$verified, ]
if (nrow(problematic) > 0) {
  cat("Warning: These foils are too similar to target:\n")
  print(problematic[, c("foil_path", "similarity")])
}

# Check similarity distribution
cat("Mean similarity:", round(mean(lineup_result$similarity), 3), "\n")
cat("SD similarity:", round(sd(lineup_result$similarity), 3), "\n")
cat("Range:", range(lineup_result$similarity), "\n")
```

### Fairness Assessment

A fair lineup should have:
- **Low mean similarity** - Foils sufficiently different from target
- **Low variance** - All foils similarly dissimilar (homogeneous)
- **No verified matches** - No foil exceeds same-person threshold

```{r fairness-assessment, purl=FALSE}
# Assess lineup fairness
mean_sim <- mean(lineup_result$similarity)
sd_sim <- sd(lineup_result$similarity)
n_verified <- sum(lineup_result$verified)

cat("Lineup Fairness Assessment:\n")
cat("Mean similarity:", round(mean_sim, 3), "\n")
cat("Similarity SD:", round(sd_sim, 3), "\n")
cat("Verified as same:", n_verified, "\n\n")

if (mean_sim > 0.7) {
  cat("WARNING: High mean similarity - lineup may be biased toward target\n")
} else if (mean_sim < 0.4) {
  cat("CAUTION: Very low similarity - foils may not match description\n")
} else {
  cat("OK: Similarity in acceptable range\n")
}

if (sd_sim > 0.15) {
  cat("WARNING: High variability - lineup not homogeneous\n")
}

if (n_verified > 0) {
  cat("WARNING:", n_verified, "foil(s) verified as same person\n")
}
```

## Models and Metrics

Different models and metrics serve different purposes.

### Available Models

```{r models, purl=FALSE}
# List all models
available_models()

# Models in order of recommendation for lineup research:
# 1. ArcFace    - State-of-the-art, best overall performance
# 2. Facenet512 - Very good, slightly faster than ArcFace
# 3. Facenet    - Good, faster inference
# 4. VGG-Face   - Older but reliable baseline
# 5. OpenFace   - Fastest, lower accuracy
```

### Distance Metrics

```{r metrics, purl=FALSE}
# Three metrics available:

# 1. Cosine (RECOMMENDED for most cases)
#    - Range: [0, 1] (can convert to similarity)
#    - Focuses on angle between embeddings
#    - Most commonly used in face recognition

# 2. Euclidean L2 (normalized Euclidean)
#    - Normalized vector distance
#    - Good for comparing across models

# 3. Euclidean
#    - Raw vector distance
#    - Less commonly used
```

### Comparing Models

```{r model-comparison, purl=FALSE}
# Compare different models on same lineup
models_to_test <- c("ArcFace", "Facenet512", "VGG-Face")

comparison_results <- list()

for (model in models_to_test) {
  result <- lineup_similarity(
    target_path = "path/to/target.jpg",
    foil_paths = foil_paths,
    model = model,
    metric = "cosine"
  )

  comparison_results[[model]] <- data.frame(
    model = model,
    mean_similarity = mean(result$similarity),
    sd_similarity = sd(result$similarity),
    n_verified = sum(result$verified)
  )
}

comparison_df <- do.call(rbind, comparison_results)
print(comparison_df)
```

**Interpretation:**
- Models may give different absolute scores
- But relative ordering (which foils are most similar) usually consistent
- Use same model consistently within a study
- ArcFace generally most accurate

## Visualization

### Display Lineup with Similarity Scores

```{r display-lineup, purl=FALSE}
# Create visual display of lineup
display_lineup(
  target_path = "path/to/target.jpg",
  foil_paths = foil_paths,
  similarity_scores = lineup_result$similarity,
  labels = paste("Foil", 1:5),
  title = "Lineup Similarity Analysis"
)

# Save the visualization
ggsave("lineup_display.png", width = 10, height = 8)
```

### Similarity Heatmap

```{r similarity-heatmap, purl=FALSE}
# For multiple lineups, create heatmap
library(ggplot2)

# Suppose we have analyzed multiple lineups
lineup_ids <- c("Lineup_1", "Lineup_2", "Lineup_3")
foil_positions <- 1:5

# Create example data (replace with your actual data)
similarity_matrix <- matrix(
  runif(15, 0.3, 0.8),  # Example data
  nrow = 3,
  ncol = 5,
  dimnames = list(lineup_ids, paste("Foil", foil_positions))
)

# Convert to long format for ggplot
library(reshape2)
sim_long <- melt(similarity_matrix)
colnames(sim_long) <- c("Lineup", "Foil", "Similarity")

# Create heatmap
ggplot(sim_long, aes(x = Foil, y = Lineup, fill = Similarity)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "yellow", high = "red",
    midpoint = 0.6,
    limits = c(0, 1)
  ) +
  geom_text(aes(label = round(Similarity, 2)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "Lineup Similarity Heatmap",
       subtitle = "Color indicates target-foil similarity")
```

### Similarity Distribution

```{r sim-distribution, purl=FALSE}
# Plot distribution of similarities
ggplot(lineup_result, aes(x = similarity)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  geom_vline(aes(xintercept = threshold),
             color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = threshold, y = Inf,
           label = "Verification\nThreshold",
           vjust = 1.5, color = "red") +
  labs(
    title = "Distribution of Target-Foil Similarity",
    x = "Similarity Score",
    y = "Count"
  ) +
  theme_minimal()
```

## Advanced Features

### Extract Face Embeddings

Embeddings are the deep learning representations of faces. They can be used for advanced analyses.

```{r embeddings, purl=FALSE}
# Get embedding for a face
embedding <- get_embedding(
  img_path = "path/to/face.jpg",
  model = "ArcFace"
)

# Embedding is a numeric vector (typically 512 dimensions)
length(embedding)  # 512 for ArcFace

# Get embeddings for comparison
result_with_emb <- face_similarity(
  img1_path = "path/to/target.jpg",
  img2_path = "path/to/foil.jpg",
  model = "ArcFace",
  return_embeddings = TRUE
)

target_embedding <- result_with_emb$embeddings$img1
foil_embedding <- result_with_emb$embeddings$img2
```

### Dimensionality Reduction Visualization

Visualize lineup in 2D space using embeddings:

```{r tsne-viz, purl=FALSE}
# Requires Rtsne package
library(Rtsne)

# Collect embeddings for all lineup members
all_embeddings <- list()

# Target embedding
all_embeddings[["Target"]] <- get_embedding("path/to/target.jpg", model = "ArcFace")

# Foil embeddings
for (i in seq_along(foil_paths)) {
  all_embeddings[[paste0("Foil_", i)]] <- get_embedding(foil_paths[i], model = "ArcFace")
}

# Convert to matrix
embedding_matrix <- do.call(rbind, all_embeddings)

# Run t-SNE
set.seed(42)
tsne_result <- Rtsne(embedding_matrix, dims = 2, perplexity = 2)

# Create visualization dataframe
viz_df <- data.frame(
  x = tsne_result$Y[, 1],
  y = tsne_result$Y[, 2],
  member = names(all_embeddings),
  type = ifelse(names(all_embeddings) == "Target", "Target", "Foil")
)

# Plot
ggplot(viz_df, aes(x = x, y = y, color = type, label = member)) +
  geom_point(size = 4) +
  geom_text(vjust = -1) +
  scale_color_manual(values = c("Target" = "red", "Foil" = "blue")) +
  labs(
    title = "Lineup Members in Embedding Space",
    subtitle = "t-SNE projection of face embeddings",
    x = "t-SNE Dimension 1",
    y = "t-SNE Dimension 2"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Batch Processing Multiple Lineups

```{r batch-processing, purl=FALSE}
# Process multiple lineups efficiently
lineup_list <- list(
  lineup1 = list(target = "target1.jpg", foils = c("1a.jpg", "1b.jpg", "1c.jpg")),
  lineup2 = list(target = "target2.jpg", foils = c("2a.jpg", "2b.jpg", "2c.jpg")),
  lineup3 = list(target = "target3.jpg", foils = c("3a.jpg", "3b.jpg", "3c.jpg"))
)

# Analyze all lineups
results_list <- lapply(names(lineup_list), function(lineup_name) {
  lineup <- lineup_list[[lineup_name]]

  result <- lineup_similarity(
    target_path = lineup$target,
    foil_paths = lineup$foils,
    model = "ArcFace"
  )

  # Add lineup identifier
  result$lineup_id <- lineup_name

  result
})

# Combine results
all_results <- do.call(rbind, results_list)

# Summary by lineup
library(dplyr)
lineup_summary <- all_results %>%
  group_by(lineup_id) %>%
  summarise(
    mean_similarity = mean(similarity),
    sd_similarity = sd(similarity),
    min_similarity = min(similarity),
    max_similarity = max(similarity),
    n_verified = sum(verified)
  )

print(lineup_summary)
```

## Integration with Traditional Measures

Combine deep learning similarity with traditional lineup fairness measures.

### Compare with Effective Size

```{r compare-effective-size, purl=FALSE}
# Traditional effective size from mock witnesses
# (Using example data)
mock_choices <- c(3, 2, 5, 6, 1, 3, 3, 5, 6, 4)  # Position 3 is suspect
trad_eff_size <- esize_T(mock_choices, pos = 3, k = 6)

cat("Traditional effective size:", round(trad_eff_size, 2), "\n")

# Face similarity metric
# Lower mean similarity = more "effective" lineup members
mean_similarity <- mean(lineup_result$similarity)

cat("Mean face similarity:", round(mean_similarity, 3), "\n")

# Correlation across multiple lineups
# (With sufficient data, examine relationship)
```

### Similarity-Weighted Effective Size

Create a hybrid measure:

```{r hybrid-measure, purl=FALSE}
# Weight mock witness data by face similarity
# Foils that are objectively dissimilar should be more likely chosen
# (This is a conceptual example)

# Get similarity scores
similarities <- lineup_result$similarity

# Compute "expected" selection probabilities based on similarity
# (Lower similarity = higher expected selection for fair lineup)
dissimilarity <- 1 - similarities
expected_prob <- dissimilarity / sum(dissimilarity)

# Compare to observed selections
# observed_prob <- table(mock_choices[-which(mock_choices == suspect_pos)]) /
#                  length(mock_choices[-which(mock_choices == suspect_pos)])

# Similarity-adjusted bias metric
# (Larger discrepancy = more bias after accounting for appearance)
```

### Validation Studies

Examine predictive validity:

```{r validation, purl=FALSE}
# Research question: Do objective similarity scores predict mock witness choices?

# For each lineup:
# - Get face similarity scores
# - Collect mock witness data
# - Correlate: Do more similar foils get chosen more often?

# Example correlation structure (use your data):
# cor.test(foil_similarities, foil_selection_rates)
```

## Practical Guidelines

### Interpreting Similarity Scores

**Similarity ranges (for ArcFace + cosine):**
- `> 0.80`: Very high similarity - likely lineup bias
- `0.70-0.80`: High similarity - check if acceptable for your context
- `0.55-0.70`: Moderate similarity - generally acceptable
- `0.40-0.55`: Low similarity - foils sufficiently different
- `< 0.40`: Very low similarity - foils may not match description

**Context matters:**
- Acceptable similarity depends on description specificity
- More specific descriptions allow tighter similarity ranges
- Less specific descriptions require broader ranges

### Best Practices

1. **Consistency**
   - Use same model throughout a study
   - Use same metric throughout a study
   - Use same verification thresholds

2. **Validation**
   - Compare similarity scores to mock witness data when available
   - Validate that your similarity threshold makes sense for your context

3. **Complementarity**
   - Use face similarity alongside traditional mock witness studies
   - Objective similarity + subjective judgments provide comprehensive assessment

4. **Documentation**
   - Report model and metric used
   - Report mean and SD of similarities
   - Report verification threshold and number verified

5. **Image Quality**
   - Use high-quality, front-facing photos
   - Ensure good lighting and resolution
   - Consistent photo conditions across lineup members

### Troubleshooting

**"No face detected"**
- Image quality too poor
- Face not front-facing enough
- Try different detector: `detector = "opencv"` is more lenient

**Embeddings fail**
- Check image file paths are correct
- Ensure images are standard formats (jpg, png)
- Verify Python environment is activated

**Inconsistent results**
- Deep learning models are deterministic
- If results change, check if images were modified
- Verify same model/metric used

**Very high/low similarities**
- Check image quality
- Verify correct target-foil pairing
- Consider if lineup truly is biased/fair

## Example Workflow

Complete analysis from start to finish:

```{r complete-workflow, purl=FALSE}
# 1. Setup (first time only)
# install_r4lineups_python()

# 2. Load package
library(r4lineups)

# 3. Analyze lineup
lineup_sim <- lineup_similarity(
  target_path = "suspect_photo.jpg",
  foil_paths = c("foil1.jpg", "foil2.jpg", "foil3.jpg",
                 "foil4.jpg", "foil5.jpg"),
  model = "ArcFace",
  metric = "cosine"
)

# 4. Assess fairness
mean_sim <- mean(lineup_sim$similarity)
sd_sim <- sd(lineup_sim$similarity)
n_verified <- sum(lineup_sim$verified)

# 5. Generate report
cat("=== Lineup Fairness Report ===\n\n")
cat("Objective Similarity Analysis\n")
cat("Model: ArcFace | Metric: Cosine\n\n")
cat("Results:\n")
cat("  Mean similarity:", round(mean_sim, 3), "\n")
cat("  SD similarity:", round(sd_sim, 3), "\n")
cat("  Range:", round(min(lineup_sim$similarity), 3), "-",
    round(max(lineup_sim$similarity), 3), "\n")
cat("  Verified as same person:", n_verified, "of 5 foils\n\n")

if (mean_sim > 0.7 || n_verified > 0) {
  cat("CONCERN: High similarity indicates potential lineup bias\n")
} else if (mean_sim < 0.55) {
  cat("ACCEPTABLE: Foils sufficiently dissimilar from target\n")
} else {
  cat("BORDERLINE: Consider additional assessment\n")
}

# 6. Visualize
display_lineup(
  target_path = "suspect_photo.jpg",
  foil_paths = c("foil1.jpg", "foil2.jpg", "foil3.jpg",
                 "foil4.jpg", "foil5.jpg"),
  similarity_scores = lineup_sim$similarity,
  title = "Lineup 001 - Similarity Analysis"
)

# 7. Save results
write.csv(lineup_sim, "lineup_001_similarity.csv", row.names = FALSE)
```

## References

**Deep Learning Face Recognition:**
- Deng, J., Guo, J., Xue, N., & Zafeiriou, S. (2019). ArcFace: Additive angular margin loss for deep face recognition. *CVPR 2019*.

- Schroff, F., Kalenichenko, D., & Philbin, J. (2015). FaceNet: A unified embedding for face recognition and clustering. *CVPR 2015*.

**DeepFace Library:**
- Serengil, S. I., & Ozpinar, A. (2020). LightFace: A hybrid deep face recognition framework. *ASYU 2020*.

**Lineup Research Applications:**
- Tredoux, C. G. (2002). A direct measure of facial similarity and its relation to human similarity perceptions. *Journal of Experimental Psychology: Applied, 8*(3), 180-193.

- Colloff, M. F., Wade, K. A., & Strange, D. (2016). Unfair lineups make witnesses more likely to confuse innocent and guilty suspects. *Psychological Science, 27*(9), 1227-1239.

## Session Info

```{r session-info, purl=FALSE}
sessionInfo()

# Check Python environment
reticulate::py_config()
```
