---
title: "RAC Analysis: Response Time-Accuracy Characteristics"
author: "Colin Tredoux"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{RAC Analysis: Response Time-Accuracy Characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette demonstrates **RAC (Response Time-Accuracy Characteristic) analysis** for eyewitness lineup data, a new feature in r4lineups inspired by pyWitness (Mickes et al., 2024).

## What is RAC Analysis?

RAC analysis examines the relationship between **response time** and **accuracy** in lineup identifications. It plots suspect ID accuracy as a function of response time, similar to how CAC analysis plots accuracy as a function of confidence.

## Why Use RAC?

Response time is an important variable in eyewitness research because:

1. **Fast responses often indicate strong memory** - Witnesses with strong memory of the culprit can identify them quickly
2. **Response time is objective** - Unlike confidence, it cannot be influenced by post-event information or verbal overshadowing
3. **Complementary to confidence** - RAC and CAC together provide a fuller picture of memory strength
4. **Forensically useful** - Response time can be recorded automatically in video-based lineups

## Key References

- **Seale-Carlisle et al. (2019)**. Confidence and response time as indicators of eyewitness identification accuracy in the lab and in the real world. *Journal of Applied Research in Memory and Cognition, 8*(4), 420-428.

- **Mickes et al. (2024)**. pyWitness 1.0: A python eyewitness identification analysis toolkit. *Behavior Research Methods, 56*, 1533-1550.

# Setup

```{r load_package}
library(r4lineups)
library(ggplot2)
```

# Simulating Data with Response Times

Since most existing datasets don't include response times, we'll simulate realistic data:

```{r simulate_data}
set.seed(123)

# Simulate 500 lineup trials
# Target-present trials
tp_data <- data.frame(
  target_present = TRUE,
  # 200 correct IDs (fast, mean = 7s)
  identification = c(rep("suspect", 200), rep("filler", 40), rep("reject", 60)),
  response_time = c(
    rnorm(200, mean = 7000, sd = 2500),  # Correct IDs: fast
    rnorm(40, mean = 11000, sd = 3000),   # Filler picks: slower
    rnorm(60, mean = 14000, sd = 4000)    # Rejections: slowest
  )
)

# Target-absent trials
ta_data <- data.frame(
  target_present = FALSE,
  # Innocent suspect picks are slower (weaker memory)
  identification = c(rep("suspect", 50), rep("filler", 100), rep("reject", 50)),
  response_time = c(
    rnorm(50, mean = 12000, sd = 3500),   # False IDs: slow
    rnorm(100, mean = 10000, sd = 3000),  # Filler picks
    rnorm(50, mean = 13000, sd = 4000)    # Correct rejections
  )
)

# Combine and ensure positive times
lineup_data <- rbind(tp_data, ta_data)
lineup_data$response_time <- pmax(1000, lineup_data$response_time)

# Shuffle rows
lineup_data <- lineup_data[sample(nrow(lineup_data)), ]

# Summary
table(lineup_data$target_present, lineup_data$identification)
summary(lineup_data$response_time)
```

# Basic RAC Analysis

## Define Time Bins

Response time is typically continuous, so we need to bin it:

```{r time_bins}
# Define bins in milliseconds (0-5s, 5-10s, 10-15s, 15-20s, 20s+)
time_bins <- c(0, 5000, 10000, 15000, 20000, 99999)
```

## Compute RAC

```{r compute_rac}
rac_result <- make_rac(
  data = lineup_data,
  lineup_size = 6,
  time_bins = time_bins,
  show_plot = FALSE
)

# Print results
print(rac_result)
```

## Interpret the Results

The RAC data shows:

- **Accuracy** decreases as response time increases
- **Fastest responses** (0-5s): ~`r round(rac_result$rac_data$accuracy[1], 2)` accuracy
- **Slowest responses** (15-20s): ~`r round(rac_result$rac_data$accuracy[4], 2)` accuracy

This pattern is consistent with the "strong memory = fast and accurate" principle.

## Visualize the RAC Curve

```{r plot_rac, fig.width=7, fig.height=5}
# Display the plot
print(rac_result$plot + theme_minimal(base_size = 12))
```

# RAC vs CAC Comparison

Let's compare RAC with CAC analysis using data where confidence and response time are correlated.

## Add Confidence Ratings

```{r add_confidence}
# Simulate confidence negatively correlated with response time
# Fast = high confidence, Slow = low confidence
lineup_data$confidence <- 100 - (lineup_data$response_time / 200) +
                          rnorm(nrow(lineup_data), 0, 10)
lineup_data$confidence <- pmin(100, pmax(0, lineup_data$confidence))

# Check correlation
cor(lineup_data$response_time, lineup_data$confidence)
```

## Compute Both Analyses

```{r rac_vs_cac}
# RAC analysis
rac <- make_rac(
  data = lineup_data,
  time_bins = time_bins,
  show_plot = FALSE
)

# CAC analysis
cac <- make_cac(
  data = lineup_data,
  confidence_bins = c(0, 60, 80, 100),
  show_plot = FALSE
)

# Display both
print(rac$plot + ggtitle("RAC: Response Time-Accuracy"))
print(cac$plot + ggtitle("CAC: Confidence-Accuracy"))
```

## Key Differences

| Feature | RAC | CAC |
|---------|-----|-----|
| **X-axis** | Response time | Confidence |
| **Measure** | Objective (automatic) | Subjective (self-report) |
| **Interpretation** | Faster = stronger memory | Higher confidence = stronger memory |
| **Robustness** | Not affected by verbal overshadowing | Can be influenced by post-event info |
| **Use case** | Video lineups, lab studies | Court testimony, field studies |

# Fine-Grained Analysis

Using finer time bins reveals more detail:

```{r fine_bins, fig.width=7, fig.height=5}
# Create bins every 2.5 seconds
fine_bins <- seq(0, 25000, by = 2500)

rac_fine <- make_rac(
  data = lineup_data,
  time_bins = fine_bins,
  show_plot = TRUE,
  show_errorbars = TRUE,
  show_n = TRUE
)

# Note: Check sample sizes (n) - some bins may be small
print(rac_fine$rac_data[, c("response_time", "n_total", "accuracy")])
```

**Warning**: Finer bins provide more detail but reduce sample sizes per bin. Always check the `n_total` column to ensure adequate samples.

# Practical Applications

## 1. Identifying Optimal Decision Time

RAC analysis can help identify response time thresholds for high accuracy:

```{r threshold}
# Find time bins with accuracy > 0.80
high_accuracy_bins <- rac_result$rac_data[rac_result$rac_data$accuracy > 0.80, ]
print(high_accuracy_bins[, c("response_time", "mean_time", "accuracy", "n_total")])
```

## 2. Comparing Conditions

RAC can compare different experimental conditions:

```{r compare_conditions, eval=FALSE}
# Example structure (not run)
condition1_rac <- make_rac(data_condition1, time_bins = time_bins)
condition2_rac <- make_rac(data_condition2, time_bins = time_bins)

# Compare visually or statistically
# (bootstrap methods for significance testing)
```

## 3. Custom Analysis

Access the raw RAC data for custom analyses:

```{r custom_analysis}
# Extract data
rac_data <- rac_result$rac_data

# Custom plot with smoothing
ggplot(rac_data, aes(x = mean_time / 1000, y = accuracy)) +
  geom_point(aes(size = n_total), color = "darkblue", alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  theme_bw(base_size = 14) +
  labs(
    x = "Mean Response Time (seconds)",
    y = "Accuracy",
    title = "RAC with LOESS Smoothing",
    size = "N Suspect IDs"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = 0.5, linetype = "dashed")
```

# Technical Details

## How RAC is Computed

For each response time bin:

$$\text{Accuracy} = \frac{\text{Correct Suspect IDs}}{\text{Correct Suspect IDs} + \text{Incorrect Suspect IDs}}$$

Where:
- **Correct Suspect IDs**: Target-present lineups where suspect was identified
- **Incorrect Suspect IDs**: Target-absent lineups where suspect was identified + (filler IDs / lineup size)

## Standard Errors

Standard errors are computed assuming binomial distribution:

$$SE = \sqrt{\frac{p(1-p)}{n}}$$

where $p$ is the accuracy and $n$ is the total number of suspect IDs.

# Real Data Considerations

When using RAC with real eyewitness data:

1. **Time Units**: Ensure response times are in the same units (ms or seconds)

2. **Binning Strategy**:
   - Too few bins → loss of detail
   - Too many bins → small samples, high variance
   - Recommended: 4-6 bins

3. **Extreme Values**: Consider removing or capping extreme outliers

4. **Missing Data**: Handle trials without response times appropriately

5. **Sequential Lineups**: Response time may have different meaning than in simultaneous lineups

# Example with Real Data Structure

```{r real_data_structure, eval=FALSE}
# Your data should look like this:
my_data <- data.frame(
  target_present = c(TRUE, FALSE, TRUE, ...),   # Logical
  identification = c("suspect", "filler", ...),  # Character
  response_time = c(8234, 12456, 6789, ...)     # Numeric (ms)
)

# Run analysis
result <- make_rac(
  data = my_data,
  lineup_size = 6,
  time_bins = c(0, 5000, 10000, 15000, 20000, 30000),
  time_units = "ms"
)

# View
print(result)
result$plot
```

# Exporting RAC Data

Export for use in other software or publications:

```{r export, eval=FALSE}
# As CSV
write.csv(rac_result$rac_data, "rac_analysis.csv", row.names = FALSE)

# As table for manuscript
library(knitr)
kable(rac_result$rac_data, digits = 3, caption = "RAC Analysis Results")
```

# Summary

RAC analysis is a valuable tool for eyewitness research:

✓ **Complements CAC analysis** - Objective measure of memory strength
✓ **Easy to implement** - Simple function call with time bins
✓ **Robust to bias** - Not influenced by verbal overshadowing
✓ **Forensically relevant** - Applicable to video lineups

## Recommended Workflow

1. Collect response time data during lineup administration
2. Choose appropriate time bins (4-6 bins recommended)
3. Run `make_rac()` with your data
4. Interpret the speed-accuracy relationship
5. Compare with CAC analysis
6. Report both in publications for completeness

## Next Steps

- Explore **CAC analysis** with `make_cac()`
- Try **ROC analysis** with `make_roc()` or `make_fullroc()`
- Conduct **bootstrap comparisons** between conditions
- Examine **individual differences** in response time patterns

# References

1. Seale-Carlisle, T. M., Colloff, M. F., Flowe, H. D., Wells, W., Wixted, J. T., & Mickes, L. (2019). Confidence and response time as indicators of eyewitness identification accuracy in the lab and in the real world. *Journal of Applied Research in Memory and Cognition, 8*(4), 420-428.

2. Mickes, L., Seale-Carlisle, T. M., Chen, X., & Boogert, S. (2024). pyWitness 1.0: A python eyewitness identification analysis toolkit. *Behavior Research Methods, 56*, 1533-1550.

3. Mickes, L. (2015). Receiver operating characteristic analysis and confidence-accuracy characteristic analysis in investigations of system variables and estimator variables that affect eyewitness memory. *Journal of Applied Research in Memory and Cognition, 4*(2), 93-102.

4. Wixted, J. T., & Wells, G. L. (2017). The relationship between eyewitness confidence and identification accuracy: A new synthesis. *Psychological Science in the Public Interest, 18*(1), 10-65.

---

*This vignette was created for r4lineups version `r packageVersion("r4lineups")`*
