---
title: "SDT via GLM/GLMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SDT via GLM/GLMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

This vignette shows how to estimate signal‑detection metrics from **old/new recognition**
data using GLM/GLMM methods, following the approach described in Wright, Horry, & Skagerberg (2009).

Key idea:

- Encode item status (`is_old`) and response (`said_old`) as 0/1.
- Fit a **probit** or **logit** GLM/GLMM.
- The coefficient on `is_old` corresponds to **d′** (probit) or **lnOR** (logit).
- With centered coding (`is_old` → -0.5/+0.5), the intercept corresponds to the **criterion**.

# Simulated data

```{r}
library(r4lineups)

set.seed(123)
n_subj <- 40
n_items <- 60
df <- data.frame(
  subject_id = rep(1:n_subj, each = n_items),
  item_id = rep(1:n_items, times = n_subj)
)
df$is_old <- rbinom(nrow(df), 1, 0.5)

# true d' around 1.2 with mild subject variability
subj_shift <- rnorm(n_subj, 0, 0.2)
linpred <- (df$is_old * 1.2) + subj_shift[df$subject_id]
df$said_old <- rbinom(nrow(df), 1, stats::pnorm(linpred))
```

# Single‑level GLM (probit)

```{r}
glm_fit <- fit_sdt_glm(df, is_old = "is_old", said_old = "said_old", link = "probit")
extract_sdt_metrics(glm_fit)
```

# Multilevel GLMM (random intercepts)

```{r}
glmm_fit <- fit_sdt_glmm(df,
                         is_old = "is_old",
                         said_old = "said_old",
                         subject_id = "subject_id",
                         item_id = "item_id",
                         link = "probit",
                         random_slope = FALSE)
extract_sdt_metrics(glmm_fit)
```

# Including covariates (optional)

```{r}
df$delay <- rnorm(nrow(df), 0, 1)
glm_fit_cov <- fit_sdt_glm(df,
                           is_old = "is_old",
                           said_old = "said_old",
                           covariates = "delay",
                           link = "probit",
                           interactions = TRUE)
summary(glm_fit_cov)
```

# Notes

- **Probit link** ⇒ coefficient on `is_old` = d′.  
- **Logit link** ⇒ coefficient on `is_old` = ln(OR).  
- Criterion is returned only when `is_old` is centered (default).

For multilevel models, `fit_sdt_glmm()` requires **lme4**.
