---
title: "Data Simulation and Power Analysis for Lineup Studies"
author: "r4lineups package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Simulation and Power Analysis for Lineup Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates how to use r4lineups' data simulation and power analysis tools for planning eyewitness identification studies. These tools help researchers:

- **Plan studies**: Determine required sample sizes for detecting effects
- **Validate methods**: Test if analyses recover known parameters
- **Explore scenarios**: Compare different experimental designs
- **Teach concepts**: Demonstrate signal detection theory principles

The simulation framework implements a Signal Detection Theory (SDT) model with the MAX decision rule, following the methodology used in eyewitness identification research (Wixted et al., 2018).

## Signal Detection Model

The simulation generates data using the following SDT framework:

**Memory strength distributions:**
- Targets (guilty suspects): Normal(d', 1)
- Lures (foils/innocent suspects): Normal(0, 1)

**MAX decision rule:**
- The witness selects the lineup member with the highest memory strength
- If all memory strengths are below criterion c, the lineup is rejected

**Parameters:**
- `d_prime`: Discriminability between targets and lures (higher = better memory)
- `c_criterion`: Decision criterion (higher = more conservative)
- `lineup_size`: Number of lineup members
- `conf_levels`: Number of confidence scale points

## Basic Data Simulation

### Simulating a Single Dataset

```{r basic_simulation}
library(r4lineups)

# Set seed for reproducibility
set.seed(2026)

# Simulate lineup data with moderate discriminability
sim_data <- simulate_lineup_data(
  n_tp = 200,           # 200 target-present lineups
  n_ta = 200,           # 200 target-absent lineups
  d_prime = 1.5,        # Moderate discriminability
  c_criterion = 0.5,    # Neutral criterion
  lineup_size = 6,      # Standard 6-person lineup
  conf_levels = 5       # 5-point confidence scale
)

# View structure
head(sim_data)

# Summary
table(sim_data$target_present, sim_data$identification)
```

The simulated data has the standard format required by r4lineups functions:
- `target_present`: TRUE (guilty suspect) or FALSE (innocent suspect)
- `identification`: "suspect", "filler", or "reject"
- `confidence`: Confidence rating (1 to conf_levels)
- `response_time`: Simulated response time (optional)

### Exploring Discriminability Levels

```{r discriminability_levels}
# Weak discriminability
weak_data <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 1.0,
  lineup_size = 6
)

# Strong discriminability
strong_data <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 2.5,
  lineup_size = 6
)

# Compare suspect ID rates
cat("Weak d' = 1.0:\n")
cat("  TP suspect IDs:",
    mean(weak_data$target_present & weak_data$identification == "suspect"), "\n")
cat("  TA suspect IDs:",
    mean(!weak_data$target_present & weak_data$identification == "suspect"), "\n\n")

cat("Strong d' = 2.5:\n")
cat("  TP suspect IDs:",
    mean(strong_data$target_present & strong_data$identification == "suspect"), "\n")
cat("  TA suspect IDs:",
    mean(!strong_data$target_present & strong_data$identification == "suspect"), "\n")
```

As expected, higher discriminability leads to:
- More correct IDs (target-present suspect IDs)
- Fewer false IDs (target-absent suspect IDs)

## Power Analysis for Study Planning

### Simple Power Analysis

Use `simulate_power_analysis()` to determine the sample size needed to detect an effect:

```{r power_analysis_basic}
# Power analysis for detecting d' = 1.5 with ROC pAUC
power_result <- simulate_power_analysis(
  sample_sizes = c(50, 100, 150, 200, 300),
  d_prime = 1.5,
  n_simulations = 100,  # Use 500+ for real studies
  alpha = 0.05
)

# View results
print(power_result)

# Plot power curve
plot(power_result)
```

The power analysis shows how power increases with sample size. For 80% power to detect d' = 1.5, we typically need 150-200 participants per condition.

### Different Discriminability Levels

Power analysis for different discriminability levels:

```{r power_comparison}
# Power to detect d' = 2.0
power_high <- simulate_power_analysis(
  sample_sizes = c(100, 150, 200, 250, 300),
  d_prime = 2.0,
  n_simulations = 100
)

print(power_high)
plot(power_high)
```

The plot shows that with higher discriminability (d' = 2.0), smaller sample sizes are needed to achieve adequate power.

## Integration with r4lineups Analyses

Simulated data works seamlessly with all r4lineups functions:

### ROC Analysis

```{r roc_analysis}
# Generate data
sim_roc <- simulate_lineup_data(
  n_tp = 200, n_ta = 200,
  d_prime = 1.8,
  conf_levels = 5
)

# Compute ROC curve
roc_result <- make_roc(sim_roc, lineup_size = 6)
print(roc_result)
```

### CAC Analysis

```{r cac_analysis}
# Confidence-Accuracy Characteristic
cac_result <- make_cac(sim_roc)
print(cac_result)
```

### RAC Analysis

```{r rac_analysis}
# Generate data with response times
sim_rac <- simulate_lineup_data(
  n_tp = 200, n_ta = 200,
  d_prime = 1.5,
  include_response_time = TRUE
)

# Response Time-Accuracy Characteristic
rac_result <- make_rac(
  sim_rac,
  time_bins = c(0, 5000, 10000, 15000, 20000)
)
print(rac_result)
```

### Full ROC Analysis

```{r fullroc_analysis}
# Full ROC using all responses
fullroc_result <- make_fullroc(sim_roc)
print(fullroc_result)
```

## Method Validation: Parameter Recovery

Use simulation to test if your analyses can recover known parameters:

```{r parameter_recovery, eval=FALSE}
# Simulate data with known d' = 2.0
true_dprime <- 2.0
recovery_data <- simulate_lineup_data(
  n_tp = 300, n_ta = 300,
  d_prime = true_dprime,
  lineup_size = 6,
  conf_levels = 5
)

# Fit 2-HT model and check if we recover similar discriminability
# Note: Model fitting may fail with some simulated datasets
tryCatch({
  model_2ht <- fit_winter_2ht(
    recovery_data,
    lineup_size = 6,
    target_present = "target_present",
    identification = "identification"
  )

  cat("True d':", true_dprime, "\n")
  cat("Recovered dP (detection):", round(model_2ht$parameters["dP"], 3), "\n")
  cat("Expected: Higher d' should lead to higher dP parameter\n")
}, error = function(e) {
  cat("Model fitting did not converge. Try different data or parameters.\n")
})
```

## Comparing Experimental Designs

Use simulation to compare different lineup configurations:

```{r design_comparison}
# Standard 6-person lineup
lineup_6 <- simulate_lineup_data(
  n_tp = 200, n_ta = 200,
  d_prime = 1.5,
  lineup_size = 6
)

# Smaller 4-person lineup (potentially easier)
lineup_4 <- simulate_lineup_data(
  n_tp = 200, n_ta = 200,
  d_prime = 1.5,
  lineup_size = 4
)

# Larger 8-person lineup (potentially harder)
lineup_8 <- simulate_lineup_data(
  n_tp = 200, n_ta = 200,
  d_prime = 1.5,
  lineup_size = 8
)

# Compare suspect ID rates
compare_lineups <- data.frame(
  Lineup_Size = c(4, 6, 8),
  TP_Suspect_ID = c(
    mean(lineup_4$target_present & lineup_4$identification == "suspect"),
    mean(lineup_6$target_present & lineup_6$identification == "suspect"),
    mean(lineup_8$target_present & lineup_8$identification == "suspect")
  ),
  TA_Suspect_ID = c(
    mean(!lineup_4$target_present & lineup_4$identification == "suspect"),
    mean(!lineup_6$target_present & lineup_6$identification == "suspect"),
    mean(!lineup_8$target_present & lineup_8$identification == "suspect")
  )
)

print(compare_lineups)
```

## Advanced: Custom Simulation Parameters

### Varying Decision Criterion

```{r criterion_comparison}
# Liberal criterion (low c)
liberal_data <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 1.5,
  c_criterion = 0.0  # More liberal
)

# Conservative criterion (high c)
conservative_data <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 1.5,
  c_criterion = 1.0  # More conservative
)

# Compare rejection rates
cat("Liberal criterion:\n")
cat("  Rejection rate:", mean(liberal_data$identification == "reject"), "\n\n")

cat("Conservative criterion:\n")
cat("  Rejection rate:", mean(conservative_data$identification == "reject"), "\n")
```

### Multiple Confidence Levels

```{r confidence_levels}
# 3-point scale
conf_3 <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 1.5,
  conf_levels = 3
)

# 7-point scale
conf_7 <- simulate_lineup_data(
  n_tp = 150, n_ta = 150,
  d_prime = 1.5,
  conf_levels = 7
)

cat("3-point scale: unique values =", length(unique(conf_3$confidence)), "\n")
cat("7-point scale: unique values =", length(unique(conf_7$confidence)), "\n")
```

## Practical Examples

### Example 1: Planning a Lineup Procedure Study

Research question: How many participants do we need to detect discriminability at d' = 1.5?

```{r planning_example}
# Power analysis for d' = 1.5
lineup_power <- simulate_power_analysis(
  sample_sizes = seq(50, 300, by = 25),
  d_prime = 1.5,
  n_simulations = 200,
  alpha = 0.05
)

plot(lineup_power)

# Result: We need approximately 150-200 participants per condition for 80% power
```

### Example 2: Testing a New Analysis Method

Verify that your new analysis recovers the correct underlying parameters:

```{r method_validation}
# Simulate data with known parameters
validation_data <- simulate_lineup_data(
  n_tp = 500, n_ta = 500,
  d_prime = 2.0,
  lineup_size = 6,
  conf_levels = 5,
  seed = 999
)

# Run your analysis
eig_result <- compute_eig(validation_data, prior_guilt = 0.5)

cat("EIG (Expected Information Gain):", round(eig_result$eig, 4), "bits\n")
cat("Interpretation: Higher d' should produce higher EIG\n")
```

## Best Practices

### 1. Sample Size Selection

- Use power analysis to determine minimum sample sizes
- Aim for 80% power (β = 0.20)
- Account for potential dropout/exclusions
- Consider practical constraints (time, resources)

### 2. Simulation Parameters

Choose realistic parameter values:
- **d' values**:
  - Weak: 0.8 - 1.2
  - Moderate: 1.3 - 1.8
  - Strong: 1.9 - 2.5
  - Very strong: > 2.5

- **Criterion**:
  - Liberal: 0.0 - 0.3
  - Neutral: 0.4 - 0.6
  - Conservative: 0.7 - 1.2

- **Lineup size**:
  - Typical: 6 (most common)
  - Range: 4 - 8

### 3. Number of Simulations

- Exploratory: 100-200 simulations
- Planning: 500-1000 simulations
- Publication: 1000-2000 simulations

### 4. Reproducibility

Always set a seed for reproducibility:

```{r reproducibility}
# Set seed at the start
set.seed(2026)

# Your simulation code here
reproducible_data <- simulate_lineup_data(
  n_tp = 100, n_ta = 100,
  d_prime = 1.5
)
```

## Interpretation Guidelines

### Understanding d' (Discriminability)

- **d' = 0**: No discriminability (chance performance)
- **d' = 1.0**: Weak discriminability
- **d' = 1.5**: Moderate discriminability
- **d' = 2.0**: Strong discriminability
- **d' = 2.5+**: Very strong discriminability

### Understanding c (Criterion)

- **c < 0**: Liberal (high ID rate, high false ID rate)
- **c = 0**: Neutral (balanced)
- **c > 0**: Conservative (low ID rate, low false ID rate)

## Common Pitfalls

### 1. Insufficient Sample Size

```{r pitfall_sample_size}
# Too small for reliable estimates
small_sample <- simulate_lineup_data(n_tp = 20, n_ta = 20, d_prime = 1.5)

# ROC will be unreliable
roc_small <- make_roc(small_sample, lineup_size = 6, show_plot = FALSE)
cat("Small sample pAUC:", round(roc_small$pauc, 3),
    "(unreliable with n=40)\n")
```

### 2. Unrealistic Parameter Values

```{r pitfall_parameters}
# Unrealistically high d' (rarely observed in real studies)
unrealistic <- simulate_lineup_data(
  n_tp = 100, n_ta = 100,
  d_prime = 4.0  # Too high!
)

cat("Suspect ID rate:",
    mean(unrealistic$target_present & unrealistic$identification == "suspect"),
    "\n")
cat("This is unrealistically high for real eyewitness data\n")
```

### 3. Ignoring Multiple Testing

When running many simulations, adjust for multiple comparisons:

```{r pitfall_multiple_testing}
# If testing 5 different conditions, use Bonferroni correction
alpha_corrected <- 0.05 / 5
cat("Adjusted alpha for 5 comparisons:", alpha_corrected, "\n")
```

## Summary

Key functions demonstrated:

- `simulate_lineup_data()`: Generate lineup identification data
- `simulate_power_analysis()`: Determine required sample sizes
- Integration with all r4lineups analyses (ROC, CAC, RAC, Full ROC, models)

Key takeaways:

1. Use simulation for study planning and power analysis
2. Validate methods with parameter recovery
3. Choose realistic parameter values
4. Use sufficient bootstrap/simulation samples
5. Always report your assumptions and seed values

## References

Wixted, J. T., Vul, E., Mickes, L., & Wilson, B. M. (2018). Models of lineup memory. *Cognitive Psychology, 105*, 81-114.

Mickes, L., Seale-Carlisle, T. M., Chen, X., & Boogert, S. (2024). pyWitness 1.0: A python eyewitness identification analysis toolkit. *Behavior Research Methods, 56*, 1533-1550.

Wixted, J. T., & Mickes, L. (2012). The field of eyewitness memory should abandon probative value and embrace receiver operating characteristic analysis. *Perspectives on Psychological Science, 7*(3), 275-278.

## Further Reading

- See `?simulate_lineup_data` for complete parameter documentation
- See `?simulate_power_analysis` for power analysis options
- See vignette("model_comparison") for comparing different models
- See vignette("pauc_comparison") for statistical comparison of ROC curves
